---
title: "Adding a New Entry Format"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding a New Entry Format}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Introduction

The `vitae` package provides several built-in entry formats such as `detailed_entries()`, `brief_entries()`, `honor_entries()`, and `skill_entries()`. However, you may need a custom entry type to display different information structures—for instance, a `plain_entries()` format to output a paragraph or short summary text block within a section.

This vignette documents the key architectural changes required to add a new entry type to `vitae`, using the addition of `plain_entries()` as a reference example.

---

# Architecture Overview

Adding a new entry format to `vitae` requires integrating components across LaTeX class definitions, R data structures, template format registries, and S3 print methods. 

As outlined in `R/crisp.R`, the overall workflow involves:

1. **Defining LaTeX environments and commands** in template files (`.cls` / `.tex`).
2. **Updating `new_entry_formats()`** in `R/00_entries.R` to accept the new format option.
3. **Registering the format generator function** in template definitions (e.g., `crisp_entries` in `R/crisp.R`).
4. **Creating the R constructor function** (e.g., `plain_entries()` in `R/plain.R`).
5. **Implementing S3 methods** (`tbl_sum` and `knit_print`) for the new entry class.
6. **Exporting and registering functions** in `NAMESPACE`.

---

# Step-by-Step Implementation

## 1. Define LaTeX Commands and Environments

First, the underlying LaTeX template must define how the new entry type is styled and formatted in TeX.

For the `plain` entry type in the `crisp` template (`inst/rmarkdown/templates/crisp/skeleton/crisp.cls`), the `cvtext` environment and `\cvplain` command were added:

```latex
% Define an environment for cvtext
\newenvironment{cvtext}{%
  \vspace{\acvSectionContentTopSkip}
  \vspace{1.0mm}
  \begin{center}
    \setlength\tabcolsep{1ex}
    \setlength{\extrarowheight}{0pt}
    \begin{tabular*}{\textwidth}{@{} L{\textwidth} @{}}
}{%
    \end{tabular*}
  \end{center}
}

% Define a display for plain text
% Usage: \cvplain{<text>}
\newcommand*{\cvplain}[1]{%
  \descriptionstyle{#1} \\
}
```

## 2. Update `new_entry_formats()` in `R/00_entries.R`

The `new_entry_formats()` helper function aggregates entry generator functions for a template. It must be updated to accept the new entry format parameter (`plain = NULL`):

```r
new_entry_formats <- function(brief, detailed, honor = NULL, skill = NULL, plain = NULL) {
  formats <- list(brief = brief, detailed = detailed)
  if (!is.null(honor)) formats$honor <- honor
  if (!is.null(skill)) formats$skill <- skill
  if (!is.null(plain)) formats$plain <- plain
  formats
}
```

## 3. Register LaTeX Generator in Template (`R/crisp.R`)

The generator function produces the actual LaTeX string using the commands defined in step 1. In `R/crisp.R`, the generator function for `plain` is passed to `new_entry_formats()`:

```r
crisp_entries <- new_entry_formats(
  ...
  # add a new entry with plain text, taking only one argument, the
  # text to be displayed. This is useful for adding a short
  # description or summary of a section.
  plain = function(what){
    paste(
      c(
        "\\begin{cvtext}",
        glue_alt("\t\\cvplain{<<what>>}"),
        "\\end{cvtext}"
      ),
      collapse = "\n"
    )
  },
  ...
)
```

As noted in the developer comments in `R/crisp.R`, generator functions are called by the S3 `knit_print.*` functions (e.g. `knit_print.vitae_plain()` calls `plain()`). If you define a new entry format, you register it with `set_entry_formats()` for the current template.

## 4. Construct R Entry Function (`R/plain.R`)

Create the user-facing constructor function (e.g., `plain_entries()`). This function accepts input data, evaluates expressions using `rlang`, and assigns a custom class (`vitae_plain`) inheriting from `vitae_preserve`:

```r
#' Create plain entries
#'
#' @param data A `data.frame` or `tibble` containing the plain text entries
#' @param details the column name in `data` that contains the details for each entry
#' @param .protect Logical, if TRUE (default) protects special LaTeX characters in input
#'
#' @return Object of class \code{vitae_plain}
#' @export
plain_entries <- function(data, details, .protect = TRUE) {
  plain_exprs <- list(
    details = enquo(details) %missing% NA_character_
  )

  out <- dplyr::as_tibble(map(plain_exprs, eval_tidy, data = data))
  structure(out,
    preserve = names(plain_exprs),
    protect = .protect,
    class = c("vitae_plain", "vitae_preserve", class(data))
  )
}
```

## 5. Implement S3 Methods for the Entry Class (`R/plain.R`)

### Summary Method (`tbl_sum`)

Provides a clean summary header when printing the entry object in tibble format:

```r
#' @importFrom tibble tbl_sum
#' @export
tbl_sum.vitae_plain <- function(x) {
  x <- NextMethod()
  c(x, "vitae type" = "plain entries")
}
```

### Knitr Printing Method (`knit_print`)

`knit_print.vitae_plain()` is called by `knitr` when rendering the document. It retrieves the registered `plain` format function from `entry_format_functions$format`, applies protection against special TeX characters if requested, and returns raw LaTeX via `knitr::asis_output()`:

```r
#' @importFrom knitr knit_print
#' @export
knit_print.vitae_plain <- function(x, options, ...) {
  if(is.null(entry_format_functions$format)) {
    warn("Plain entry formatter is not defined for this output format.")
    return(knit_print(tibble::as_tibble(x)))
  }

  format <- entry_format_functions$format$plain
  if (is.null(format)) {
    warning("Plain format not supported for this template, using brief format instead")
    format <- entry_format_functions$format$brief
  }

  x[is.na(x)] <- ""

  if(!(x%@%"protect")){
    protect_tex_input <- identity
  }

  knitr::asis_output(
    format(
      protect_tex_input(x$details)
    )
  )
}
```

## 6. Export and Register in `NAMESPACE`

Finally, export the new function and register S3 methods in `NAMESPACE`
by running `devtools::document()` or manually adding the following
lines to `NAMESPACE`:

```R
devtools::document()
#export(plain_entries)
#S3method(knit_print, vitae_plain)
#S3method(tbl_sum, vitae_plain)
```

---

# Usage Example

Once implemented, users can use `plain_entries()` in their R Markdown CV document as follows:

```{r eval=FALSE}
library(vitae)
library(tibble)

tribble(
  ~Summary,
  "Experienced data scientist specializing in statistical modeling, machine learning, and open-source software development."
) %>%
  plain_entries(details = Summary)
```

---

# Summary Checklist

When adding another entry format in the future, follow this process:

- [ ] Define TeX environments/commands in template `.cls` or `.tex` resource files.
- [ ] Add default parameter (e.g. `plain = NULL`) to `new_entry_formats()` in `R/00_entries.R`.
- [ ] Implement entry format function in template format definitions (e.g. `crisp_entries` in `R/crisp.R`).
- [ ] Create `R/<entry>.R` with user constructor function returning `vitae_<entry>` class.
- [ ] Implement `tbl_sum.vitae_<entry>` and `knit_print.vitae_<entry>` S3 methods.
- [ ] Export function and register S3 methods in `NAMESPACE`.
